---
title: "Pathogencity Analysis"
author: "Taylere Herrmann"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# 0.5 Set up

```{r}
setwd("/Users/taylere/Library/CloudStorage/OneDrive-Personal/R/Alternaria/Pathogenicity_Analysis")
```

```{r}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# 1. Libraries

```{r}
library(tidyverse)
library(lme4)  
library(lmerTest)
library(emmeans)  
library(multcomp)
library(multcompView)
```

# 2. Data

```{r}
assay <- read.csv("../Data/assay_analysis.csv")
assay$isolate <- trimws(assay$isolate)

assay_clean <- assay %>%
  mutate(
    isolate  = as.factor(isolate),
    crop     = as.factor(crop),
    rep      = as.factor(rep),
    cultivar = as.factor(cultivar)
  )

cat("Isolates:", nlevels(assay_clean$isolate), "\n")
cat("Crops:", levels(assay_clean$crop), "\n")
cat("Repetitions:", nlevels(assay_clean$rep), "\n")
cat("Rows:", nrow(assay_clean), "(120 x 5 x 3 = 1800, fully balanced)\n")
```

# 3. Pearson correlation

3 dpi vs 5 dpi To deermine which timepoint to use for analysis, disease
severity at 3 and 5 dpi were compared using Pearson correlation. A
strong + correlation would justify using only 5 dpi for analysis.

```{r}
cor_dpi <- cor.test(assay_clean$disease.3, assay_clean$disease.5,
                    method = "pearson")
cat("Pearson r =", round(cor_dpi$estimate, 3), "\n")
cat("p =", format(cor_dpi$p.value, scientific = TRUE), "\n")
```

Pearson correlation r = 0.71, p essentially zero (p = 9.88 x 10-276)
Strong positive correlation between disease at 3 and 5 dpi. Confirms
that ratings at 5 dpi are representative and sufficient, we don't need
both time points

# 4. Mixed Model, Isolate and Crop effects

Fixed effects: isolate, crop Random effect: (1\|rep) --\>
accounts for varation between 3 reps Response varaible: disease.5 --\>
disease severity at 5dpi

## 4.1 Full model with isolate:crop interaction

```{r}
model_full <- lmer(
  disease.5 ~ isolate + crop + isolate:crop + (1 | rep),
  data    = assay_clean,
  REML    = TRUE,
  control = lmerControl(optimizer = "bobyqa",
                        optCtrl   = list(maxfun = 2e5))
)

print(anova(model_full))
```

ANOVA with interaction (isolate:crop p=1) The interaction between
isolate and crop is non significant This means ranking of isolates by
aggressiveness was consistent regardless of which crop was tested. A
highly aggressive isolate on broccoli was also highly aggressive on kale
etc. Important biologically and justifies using the mean across all
crops

## 4.2 Simpler model without interaction

```{r}
model_no_int <- lmer(
  disease.5 ~ isolate + crop + (1 | rep),
  data    = assay_clean,
  REML    = TRUE,
  control = lmerControl(optimizer = "bobyqa",
                        optCtrl   = list(maxfun = 2e5))
)
```

## 4.3 Model comparison

```{r}
print(anova(model_no_int, model_full))
```

Model comparison: p = 0.9256 Confirms the model without the intreaction
term is just as good as the full model. Use the simpler model

```{r}
# Note: model_final is the untransformed model used only for
model_final <- model_no_int

print(anova(model_final))
```

Both isolate (p=2.2 x 10-16) and crop (p\<2.2 x 10-16) are highly sig.
Isolates sig differ in aggressiveness Crops sig. differ in
susceptibility

# 5. Residual Checks and Logit Transformation

## 5.1 Untransformed model

```{r}
residuals_model <- residuals(model_final)
sw_untransformed <- shapiro.test(residuals_model)
cat("Shapiro-Wilk (untransformed): W =", round(sw_untransformed$statistic, 4),
    "| p =", round(sw_untransformed$p.value, 4), "\n")

plot(fitted(model_final), residuals_model,
     xlab = "Fitted values", ylab = "Residuals",
     main = "Residuals vs Fitted — Untransformed Model")
abline(h = 0, col = "red", lty = 2)

qqnorm(residuals_model, main = "Q-Q Plot — Untransformed Model")
qqline(residuals_model, col = "red")
```

Shapiro-Wilk on residuals W=0.953 p = 0 residuals are not normally
distributed Residuals have diagonal banding and fan shape --\>
heteroscedasticity. In QQ plot the the points deviated substantially
from the red reference line in both tails. Distribution is to skewed
compared to normal. Model assumptions are violated

## 5.2 Logit transformation

A logit transformation may be a good fit for bounded proportion data
(Warton & Hui 2011)

```{r}
assay_clean <- assay_clean %>%
  mutate(
    disease5_logit = log((disease.5 + 0.5) / (100 - disease.5 + 0.5))
  )

# Fit same model structure on transformed data
model_logit <- lmer(
  disease5_logit ~ isolate + crop + (1 | rep),
  data    = assay_clean,
  REML    = TRUE,
  control = lmerControl(optimizer = "bobyqa",
                        optCtrl   = list(maxfun = 2e5))
)
print(anova(model_logit))
```

## 5.3 Residual checks: logit model

```{r residuals-logit}
res_logit <- residuals(model_logit)
sw_logit  <- shapiro.test(res_logit)
cat("Shapiro-Wilk (logit model): W =", round(sw_logit$statistic, 4),
    "| p =", round(sw_logit$p.value, 4), "\n")
cat("Note: with n =", nrow(assay_clean), "Shapiro-Wilk has very high power\n")
cat("and will detect trivially small deviations. Visual inspection of\n")
cat("the plots below is more informative than the p-value at this sample size.\n")

plot(fitted(model_logit), res_logit,
     xlab = "Fitted values (logit scale)", ylab = "Residuals",
     main = "Residuals vs Fitted — Logit Model")
abline(h = 0, col = "red", lty = 2)

qqnorm(res_logit, main = "Q-Q Plot — Logit Model")
qqline(res_logit, col = "red")
```

Residuals are scattered more evenly above and below zero. QQ plot, the
points follow the red line closley.

Use logit transformation for data

## 5.4 Comparison: Arcsine square root transformation

```{r}
assay_clean <- assay_clean %>%
  mutate(
    disease5_asin = asin(sqrt(disease.5 / 100))
  )

# Fit same model structure on arcsine-transformed data
model_asin <- lmer(
  disease5_asin ~ isolate + crop + (1 | rep),
  data    = assay_clean,
  REML    = TRUE,
  control = lmerControl(optimizer = "bobyqa",
                        optCtrl   = list(maxfun = 2e5))
)
```

```{r}
# Check residuals
res_asin <- residuals(model_asin)
sw_asin  <- shapiro.test(res_asin)
cat("Shapiro-Wilk (arcsine model): W =", round(sw_asin$statistic, 4),
    "| p =", round(sw_asin$p.value, 4), "\n")

plot(fitted(model_asin), res_asin,
     xlab = "Fitted values (arcsine scale)",
     ylab = "Residuals",
     main = "Residuals vs Fitted — Arcsine Model")
abline(h = 0, col = "red", lty = 2)

qqnorm(res_asin, main = "Q-Q Plot — Arcsine Model")
qqline(res_asin, col = "red")

# Compare Shapiro-Wilk W values across all three approaches
cat("\n--- Summary: Shapiro-Wilk W across approaches ---\n")
cat("Higher W = closer to normal = better\n\n")
cat("Untransformed: W =", round(sw_untransformed$statistic, 4), "\n")
cat("Logit:         W =", round(sw_logit$statistic, 4), "\n")
cat("Arcsine:       W =", round(sw_asin$statistic, 4), "\n")
```

Higher W = closer to normal = better 
Untranformed w: 0.953 
Logit:
W=0.9906 
Arcsine: W=0.9734

Logit produced the best residual behavior

# 6. Tukey HSD: Crop Means

Estimated marginal means are calculated on the logit scale and then
back-transformed to percentage scale

```{r}
emm_crop_logit   <- emmeans(model_logit, ~ crop)
tukey_crop_logit <- cld(emm_crop_logit,
                         Letters = letters,
                         adjust  = "tukey")

crop_logit_summary <- as.data.frame(tukey_crop_logit) %>%
  mutate(
    disease_pct       = exp(emmean)    / (1 + exp(emmean))    * 100,
    disease_pct_lower = exp(lower.CL)  / (1 + exp(lower.CL))  * 100,
    disease_pct_upper = exp(upper.CL)  / (1 + exp(upper.CL))  * 100,
    groups            = trimws(.group)
  )

# Back-transform SE to percentage scale using the delta method
crop_logit_summary <- crop_logit_summary %>%
  mutate(
    p_prop = exp(emmean) / (1 + exp(emmean)),
    se_pct = SE * p_prop * (1 - p_prop) * 100
  )

cat("Crop means (back-transformed to %):\n")
print(crop_logit_summary[, c("crop", "disease_pct",
                               "disease_pct_lower",
                               "disease_pct_upper", "groups")])

write.csv(crop_logit_summary, "tukey_crop_results.csv", row.names = FALSE)
cat("Saved: tukey_crop_results.csv\n")
```
Crop means (back transformed to %)

| Crop        | \% disease | \% disease lower | \% disease upper | groups |
|-------------|------------|------------------|------------------|--------|
| Cauliflower | 13.36      | 3.39             | 40.36            | a      |
| Cabbage     | 13.54      | 3.45             | 40.72            | a      |
| Kale        | 15.76      | 4.09             | 45.08            | a      |
| Collard     | 25.45      | 7.22             | 59.96            | b      |
| Broccoli    | 32.96      | 10.08            | 68.33            | c      |

# 7. Tukey HSD: Isolate Means

```{r}
emm_iso_logit   <- emmeans(model_logit, ~ isolate)
tukey_iso_logit <- cld(emm_iso_logit,
                        Letters = letters,
                        adjust  = "tukey")

isolate_logit_summary <- as.data.frame(tukey_iso_logit) %>%
  mutate(
    disease_pct               = exp(emmean)   / (1 + exp(emmean))   * 100,
    lower.CL_backtransformed  = exp(lower.CL) / (1 + exp(lower.CL)) * 100,
    upper.CL_backtransformed  = exp(upper.CL) / (1 + exp(upper.CL)) * 100,
    groups                    = trimws(.group)
  )

# Back-transform SE to percentage scale (delta method)
isolate_logit_summary <- isolate_logit_summary %>%
  mutate(
    p_prop = exp(emmean) / (1 + exp(emmean)),
    se_pct = SE * p_prop * (1 - p_prop) * 100
  )

cat("Isolate mean range (back-transformed):",
    round(min(isolate_logit_summary$disease_pct), 1), "to",
    round(max(isolate_logit_summary$disease_pct), 1), "%\n")
cat("Overall mean:",
    round(mean(isolate_logit_summary$disease_pct), 1), "%\n")

write.csv(isolate_logit_summary, "tukey_isolate_results.csv", row.names = FALSE)
```

Isolate mean range (back transformed to): 3.6 to 52.9%

Overall mean: 21.2%


# 8 Aggressiveness by county 
Kruskal Wallis bc were testing 120 isolates that arnt normally distributed 
```{r}
# County only — one row per isolate. Do NOT include year here.
iso_county <- assay_clean %>%
  dplyr::select(isolate, county) %>%
  distinct()

aggr_geo <- isolate_logit_summary %>%
  left_join(iso_county, by = "isolate")

stopifnot(nrow(aggr_geo) == 120)   # catches the duplication if it recurs

cat("--- Aggressiveness by county ---\n")
print(aggr_geo %>%
        group_by(county) %>%
        summarise(n = n(),
                  mean_pct = round(mean(disease_pct), 1),
                  .groups = "drop"))

kruskal.test(disease_pct ~ as.factor(county), data = aggr_geo)

```

p = 0.1529 ; not significant 

# 9. Mean Aggressiveness Per Isolate

Mean aggressiveness --\> mean disease severity at 5dpi averaged across
the 5 crops and 3 reps Estimated from logit mixed model and back
transformed to % scale

Using model derived estimates accounts for random effect of rep.

```{r}
mean_aggressiveness <- isolate_logit_summary %>%
  dplyr::select(isolate, disease_pct, se_pct, groups) %>%
  rename(
    mean_disease5 = disease_pct,
    SE            = se_pct,
    tukey_group   = groups
  )

write.csv(mean_aggressiveness, "mean_aggressiveness_per_isolate.csv",
          row.names = FALSE)
cat("Range:", round(min(mean_aggressiveness$mean_disease5), 1),
    "to", round(max(mean_aggressiveness$mean_disease5), 1), "%\n")
```

Mean aggressivness range: 3.6 to 52.9%

# 10. Figures

## 10.1 Isolate aggressivenss

```{r}
median_disease <- median(isolate_logit_summary$disease_pct)
cat("Median isolate disease severity:", round(median_disease, 1), "%\n")

isolate_logit_summary <- isolate_logit_summary %>%
  mutate(
    panel_group = ifelse(disease_pct >= median_disease,
                         "Higher aggressiveness",
                         "Lower aggressiveness"),
    panel_group = factor(panel_group,
                         levels = c("Higher aggressiveness",
                                    "Lower aggressiveness"))
  )

p_isolates <- ggplot(isolate_logit_summary,
                     aes(x = reorder(isolate, -disease_pct),
                         y = disease_pct)) +
  geom_bar(stat = "identity", fill = "darkgray", width = 0.7) +
  facet_wrap(~ panel_group, scales = "free_x", nrow = 2) +
  labs(
    x = "Isolates",
    y = "Disease percentage"
  ) +
  theme_classic() +
  theme(
    axis.text.x  = element_text(angle = 90, size = 7, family = "Arial"),
    strip.text.x = element_blank()
  )

print(p_isolates)
ggsave("Figure1_isolate_aggressiveness.tiff", p_isolates,
       dpi = 600, width = 6.5, height = 4,
       units = "in", compression = "lzw")
```

## 10.2 Frequency distribution of aggressiveness/disease severity

```{r }
p_freq <- ggplot(isolate_logit_summary, aes(x = disease_pct)) +
  geom_histogram(binwidth = 5, fill = "darkgray",
                 color = "white", boundary = 0) +
  scale_x_continuous(breaks = seq(0, 55, by = 10)) +
  scale_y_continuous(breaks = seq(0, 30, by = 5),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = "Disease severity (%)", y = "Number of isolates") +
  theme_classic()

print(p_freq)
ggsave("Figure2_frequency_distribution.tiff", p_freq,
       dpi = 600, width = 5, height = 4,
       units = "in", compression = "lzw")
```

## 10.3 Crop Susceptibility

Mean disease severity per crop with SE error bars and Tukey HSD
letters.

```{r}
p_crops <- ggplot(crop_logit_summary,
                  aes(x = reorder(crop, -disease_pct),
                      y = disease_pct)) +
  geom_bar(stat = "identity", width = 0.5, fill = "gray30") +
  geom_errorbar(aes(ymin = disease_pct - se_pct,
                    ymax = disease_pct + se_pct),
                width = 0.2) +
  geom_text(aes(label = groups,
                y     = disease_pct + se_pct),
            vjust = -0.5, hjust = 0.5, size = 4) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(
    x = "Crop",
    y = "Disease percentage"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 12)
  )

print(p_crops)
ggsave("Figure3_crop_susceptibility.tiff", p_crops,
       dpi = 600, width = 6, height = 4.5,
       units = "in", compression = "lzw")

```

Bars = SE; letters indicate Tukey HSD groupinbgs 